Causal Analysis Paper

### 2022-06-08 PaperAnalysis.Rmd
G-separation theory

###G-separation
G-separation or G computation is a method that enables the unbiased estimation of marginal causal effects by using counterfactuals. It thus responds to the rupture of two of the main assumptions discussed above: the missingness of counterfactuals on observational data (because it is impossible to observe participants under both treatment and non-treatment) and heterogeneity of effects ().

"Once estimated, the Q-model (a linear regression of the outcome on the exposure and covariates) is used to predict counterfactual outcomes for each observation under each exposure regimen. This is accomplished by plugging $a=1$ and then subsequently $a=0$ into the regression fit to obtain a predicted outcome under these 2 settings. The counterfactual
outcome associated with exposure intervention a can be represented as $Y_a$. Because we have a dichotomous treatment setting, the counterfactual outcomes correspond to
$Y_1$ for exposed and $Y_0$ for unexposed. Generating counterfactual outcomes for each observation in under exposed and unexposed settings furnishes the investigator with a full data set that is free of confounding under causal assumptions and resolves the missing data problem described above."[@snowden2011implementation]

>>>>

###G-separation

G-separation is a method that enables the unbiased estimation of marginal causal effects by using counterfactuals. It  responds to two of the effect identification issues encountered in observational studies: the missingness of counterfactuals (because it is impossible to observe participants under both treatment and non-treatment) and heterogeneity of effects (variation of effects across individuals).[@snowden2011implementation]

G-separation calculation has two steps: 
1. Fitting a Q-model, that is, a traditional regression of the outcome on the treatment and  relevant covariates.  In our case, we used the traditional regression fitted in the original paper as Q model:

$$
MMR=\beta_0 + \beta_1 L+\beta_2 GDP +\beta_3 P+ \beta_4 C
$$
In the formula, P and C respectively account for 3 year periods and the country specific effects. 
2. The Q-model is used to predict counterfactual outcomes for all the observations under both treatment settings (L=1 for flexible abortion laws and L=0 if not). More specifically, we plugged both treatment regimes in the the regression fit and obtained two predicted outcomes: $Y_{L=0}=Y_0$ for the unexposed and $Y_{L=1}=Y_1$ for the exposed. Both outcomes are then used to calculate the average $Y_1-Y_0$, which corresponds to the ATE.

With this intervention, the missingness problem is resolved because counterfactual outcomes for each observation are generated. G-separation thus provides a full dataset "free of confounding under causal assumptions."[@snowden2011implementation].

Furthermore, G-separation allows for the calculation of both ATT and ATU by restricting the ATE estimation to the treated and the untreated respectively[@Wang].
